\huge The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2021-05-18 Writeup/kohler_clarke_2021.tex
final changes to manuscript, abstract, etc

\begin{abstract}
Symmetries are present at many scales in images of natural scenes. \textcolor{ForestGreen}{Humans and other animals are highly sensitive to visual symmetry, and a large} literature has demonstrated contributions of symmetry to numerous domains of visual perception. The four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined in exactly 17 distinct ways. These \textit{wallpaper groups} represent the complete set of symmetries in 2D images and have recently found use in the vision science community as an ideal stimulus set for studying the perception of symmetries in textures. The goal of the current study is to provide a more comprehensive description of responses to symmetry in the human visual system, by collecting both brain imaging (Steady-State Visual Evoked Potentials measured using high-density EEG) and behavioral (symmetry detection thresholds) data using the entire set of wallpaper groups. This allows us to probe the hierarchy of complexity among wallpaper groups, in which simpler groups are subgroups of more complex ones. We find that this hierarchy is preserved almost perfectly in both behavior and brain activity: A multi-level Bayesian GLM indicates that for most of the 63 subgroup relationships, subgroups produce lower amplitude responses in visual cortex (posterior probability: $>0.95$ for 56 of 63) and require longer presentation durations to be reliably detected (posterior probability: $>0.95$ for \textcolor{ForestGreen}{48} of 63). This systematic pattern is seen only in visual cortex and only in components of the brain response known to be symmetric-specific. Our results show that representations of symmetries in the human brain are precise and rich in detail, and that this precision is reflected in behavior. These findings expand our understanding of symmetry perception, and open up new avenues for research on how fine-grained representations of regular textures contribute to natural vision.
\end{abstract}

@@

A wallpaper group is a topologically discrete group of isometries of the Euclidean plane, i.e. transformations that preserve distance \citep{RN1425}. The wallpaper groups differ in the number and kind of these transformations \textcolor{ForestGreen}{and we can uniquely refer to different groups using crystallographic notation (CITATION?). In brief, most groups are notated by $PXZ$, where $X\in\{1,2,3,4,6\}$ indicates the highest order of rotation symmetry and $Z\in\{m, g\}$ indicates whether the pattern contains reflection (m) or glide reflection (g). For example, $P4$ contains 4 fold rotation, while $P2M$ contains 2 fold rotation and one reflection (see Figure XXX). Two of the groups start with a $C$ rather than a $P$, ($CM$ and $CMM$) which indicates that the symmetries are specified relative to a cell that itself contains repetition. Full details of the naming convention and examples of the wallpaper groups can be found on \href{https://en.wikipedia.org/wiki/Wallpaper_group}{wikipedia}}.

In mathematical group theory, when the elements of one group is completely contained in another, the inner group is called a subgroup of the outer group \citep{RN1425}. The full list of subgroup relationships is listed in Section 1.4.2 of the Supplementary Material. Subgroup relationships between wallpaper groups can be distinguished by their indices. The index of a subgroup relationship is the number of cosets, i.e. the number of times the subgroup is found in the supergroup \citep{RN1425}. As an example, let us consider groups \textit{P2} and \textit{P6} \textcolor{ForestGreen}{(see Figure 1C)}. If we ignore the translations in two directions that both groups share, group \textit{P6} consists of the set of rotations \{$0^{\circ}$, $60^{\circ}$, $120^{\circ}$, $180^{\circ}$, $240^{\circ}$, $300^{\circ}$\}, in which \textit{P2} \{$0^{\circ}$, $180^{\circ}$\} is contained. \textit{P2} is thus a subgroup of \textit{P6}, and \textit{P6} can be generated by combining \textit{P2} with rotations \{$0^{\circ}$, $120^{\circ}$, $240^{\circ}$\}. Because \textit{P2} is repeated three times in \textit{P6}, \textit{P2} is a subgroup of \textit{P6} with index 3 \citep{RN1425}. The 17 wallpaper groups thus obey a hierarchy of complexity where simpler groups are subgroups of more complex ones \citep{RN1711}.

\begin{figure}[hptb]
\centering
\includegraphics[width=0.75\linewidth]{../figures/figure1.pdf}
\caption{\textcolor{ForestGreen}{Subgroup relationships with indices 2 (solid lines) and 3 (dashed line) are shown in (A). All other relationships can be infered by identifying the shortest path through the hierarchy, and multiplying the subgroup indices. For example, \textit{P1} is related to \textit{P6} through \textit{P6}\textrightarrow\textit{P3} (index 2) and \textit{P3}\textrightarrow\textit{P1} (index 3) so \textit{P1} is also a subgroup of \textit{P6} with index $3 \times 2 = 6$. We also show enlarged versions of some of the subgroup relationships involving \textit{PMM} (B, shown in blue) and \textit{P6} (C, shown in red) and highlight the symmetries within the subgroups to emphasize how the supergroup can be generated by adding additional transformations to the subgroup.}}
\end{figure}

\begin{figure}[hp]
\centering
\includegraphics[width=0.75\linewidth]{../figures/figure2.png}
\caption{Examples of each of the 16 wallpaper groups are shown in the left- and right-most column of the figures, next to the corresponding SSVEP (center-left) and psychological (center-right) data from each group. The SSVEP data are odd-harmonic-filtered cycle-average waveforms. In each cycle, a \textit{P1} exemplar was shown for the first 600 ms, followed by the original exemplar for the last 600 ms. Errorbars are standard error of the mean. Psychophysical data are presented as boxplots reflecting the distribution of display duration thresholds. The 16 groups are ordered by the strength of the SSVEP response, to highlight the range of response amplitudes.}
\label{fig:example_stimuli}
\end{figure}

@@

\begin{figure}[hp]
\centering
\includegraphics[width=0.75\linewidth]{../analysis/plots/subgroup_comp_eeg_rms.png}
\caption{Posterior distributions for the difference in mean SSVEP RMS amplitude. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 55/63 subgroup relationships have $p(\Delta|data)>0.99$.}
\label{fig:eeg_comp}
\end{figure}

\begin{figure}[hptb]
\centering
\includegraphics[width=0.75\linewidth]{../analysis/plots/subgroup_comp_psychophysical.png}
\caption{Posterior distributions for the difference in mean symmetry detection threshold durations. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 43/63 subgroup relationships have $p(\Delta|data)>0.99$.}
\label{fig:durations_comp}
\end{figure}

@@

\textcolor{ForestGreen}{A large literature exists on the \textit{Sustained Posterior Negativity} (SPN), a characteristic negative-going waveform that is known to reflect responses to symmetry and other forms of regularity and structure \citep{makin_2016}. The SPN scales with the proportion of reflection symmetry in displays that contain a mixture of symmetry and noise \cite{makin_formation_2020, palumbo_scaling_2015}, and both reflection, rotation and translation can produce a measurable SPN \cite{makin_2013}. It has recently been demonstrated that a holographic model of regularity \citep{van_der_helm_1996}, can predict both SPN amplitude \citep{makin_2016} and perceptual discrimination performance \citep{nucci_2007} for dot patterns that contain symmetry and other types of regularity. The available evidence suggests that the SPN and our SSVEP measurements are two distinct methods of isolating the same symmetry-related brain response: When observed in the time-domain, the symmetry-selective odd-harmonic responses produce similarly sustained waveforms (see Figure 2), odd-harmonic SSVEP responses can be measured for dot patterns similar to those used to measure the SPN \citep{RN1954}, and the one event-related study that has been published on the wallpaper groups produced SPN-like waveforms \citep{RN1959}. Future work should more firmly establish the connection and determine if the SPN can capture similarly precise symmetry responses as the SSVEPs presented here. It would also be worthwhile to ask if and how \textit{W} can computed for our random-noise based wallpaper textures where combinations of symmetries tile the plane.}

@@

\begin{figure}[tbp]
\centering
\includegraphics[width=0.5\linewidth]{../analysis/plots/model_roc_style.pdf}
\caption{This plot shows the proportion of subgroup relationships that satisfy $p(\Delta >0 | data) > x$. We can see that if we take $x = 0.95$ as our threshold, the subgroup relationships are preserved in 56/63 = 89\% and \textcolor{ForestGreen}{48/63 = 76\%} of the comparisons for the primary SSVEP and threshold duration datasets, receptively. This compares to the \textcolor{ForestGreen}{32/63= 51\%} and \textcolor{ForestGreen}{22/63 = 35\%} for the SSVEP control datasets.}
\label{fig:durations_rotations}
\end{figure}

>>>>

\begin{abstract}
\textcolor{ForestGreen}{Symmetries are present at many scales in natural scenes. Humans and other animals are highly sensitive to visual symmetry, and symmetry contributes to numerous domains of visual perception. The four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined into exactly 17 distinct regular textures. These \textit{wallpaper groups} represent the complete set of symmetries in 2D images. The current study seeks to provide a more comprehensive description of responses to symmetry in the human visual system, by collecting both brain imaging (Steady-State Visual Evoked Potentials measured using high-density EEG) and behavioral (symmetry detection thresholds) data using the entire set of wallpaper groups. This allows us to probe the hierarchy of complexity among wallpaper groups, in which simpler groups are subgroups of more complex ones. We find that both behavior and brain activity preserve the hierarchy almost perfectly: Subgroups consistently produce lower amplitude symmetry-specific responses in visual cortex and require longer presentation durations to be reliably detected. These findings expand our understanding of symmetry perception by showing that the human brain encodes symmetries with a high level of precision and detail. This opens new avenues for research on how fine-grained representations of regular textures contribute to natural vision.}
\end{abstract}

@@

A wallpaper group is a topologically discrete group of isometries of the Euclidean plane, i.e. transformations that preserve distance \citep{RN1425}. The wallpaper groups differ in the number and kind of these transformations \textcolor{ForestGreen}{and we can uniquely refer to different groups using crystallographic notation. In brief, most groups are notated by $PXZ$, where $X\in\{1,2,3,4,6\}$ indicates the highest order of rotation symmetry and $Z\in\{m, g\}$ indicates whether the pattern contains reflection (m) or glide reflection (g). For example, $P4$ contains rotation of order 4, while $P4MM$ contains rotation of order 4 and two reflections. By convention, many of the groups are given shortened names: for example, $P4MM$ is usually referred to as $P4M$, as the second reflection can be deduced from the presence of rotation of order 4 alongside a reflection. Two of the groups start with a $C$ rather than a $P$, ($CM$ and $CMM$) which indicates that the symmetries are specified relative to a cell that itself contains repetition. Full details of the naming convention can be found on \href{https://en.wikipedia.org/wiki/Wallpaper_group}{wikipedia} and examples of the wallpaper groups are shown in Figures \ref{fig:subgroups} and \ref{fig:summary}.}

In mathematical group theory, when the elements of one group is completely contained in another, the inner group is called a subgroup of the outer group \citep{RN1425}. The full list of subgroup relationships is listed in Section 1.4.2 of the Supplementary Material. Subgroup relationships between wallpaper groups can be distinguished by their indices. The index of a subgroup relationship is the number of cosets, i.e. the number of times the subgroup is found in the supergroup \citep{RN1425}. As an example, let us consider groups \textit{P2} and \textit{P6} \textcolor{ForestGreen}{(see Figure 1B)}. If we ignore the translations in two directions that both groups share, group \textit{P6} consists of the set of rotations \{$0^{\circ}$, $60^{\circ}$, $120^{\circ}$, $180^{\circ}$, $240^{\circ}$, $300^{\circ}$\}, in which \textit{P2} \{$0^{\circ}$, $180^{\circ}$\} is contained. \textit{P2} is thus a subgroup of \textit{P6}, and \textit{P6} can be generated by combining \textit{P2} with rotations \{$0^{\circ}$, $120^{\circ}$, $240^{\circ}$\}. Because \textit{P2} is repeated three times in \textit{P6}, \textit{P2} is a subgroup of \textit{P6} with index 3 \citep{RN1425}. \textcolor{ForestGreen}{Similarly, \textit{PMM} contains two reflections and rotations \{$0^{\circ}$, $180^{\circ}$\}. \textit{PMM} can be generated by adding an additional reflection to both \textit{P2} (\{$0^{\circ}$, $180^{\circ}$\}) and \textit{PM} (one reflection), so \textit{P2} and \textit{PM} are both subgroups of \textit{PMM} with index 2 (see Figure 1C).} The 17 wallpaper groups thus obey a hierarchy of complexity where simpler groups are subgroups of more complex ones \citep{RN1711}.

\begin{figure}[hptb]
\centering
\includegraphics[width=0.75\linewidth]{../figures/figure1.pdf}
\caption{\textcolor{ForestGreen}{Subgroup relationships with indices 2 (solid lines) and 3 (dashed line) are shown in (A). All other relationships can be inferred by identifying the shortest path through the hierarchy, and multiplying the subgroup indices. For example, \textit{P1} is related to \textit{P6} through \textit{P6}\textrightarrow\textit{P3} (index 2) and \textit{P3}\textrightarrow\textit{P1} (index 3) so \textit{P1} is also a subgroup of \textit{P6} with index $3 \times 2 = 6$. We also show enlarged versions of some of the subgroup relationships involving \textit{P6} (B, shown in red) and \textit{PMM} (C, shown in blue) and highlight the symmetries within the subgroups to emphasize how the supergroup can be generated by adding additional transformations to the subgroup.}}
\label{fig:subgroups}
\end{figure}

\begin{figure}[hp]
\centering
\includegraphics[width=0.75\linewidth]{../figures/figure2.png}
\caption{Examples of each of the 16 wallpaper groups are shown in the left- and right-most column of the figures, next to the corresponding SSVEP (center-left) and psychological (center-right) data from each group. The SSVEP data are odd-harmonic-filtered cycle-average waveforms. In each cycle, a \textit{P1} exemplar was shown for the first 600 ms, followed by the original exemplar for the last 600 ms. Errorbars are standard error of the mean. Psychophysical data are presented as boxplots reflecting the distribution of display duration thresholds. The 16 groups are ordered by the strength of the SSVEP response, to highlight the range of response amplitudes.}
\label{fig:summary}
\end{figure}

@@

\begin{figure}[hp]
\centering
\includegraphics[width=0.75\linewidth]{../analysis/plots/subgroup_comp_eeg_rms.png}
\caption{Posterior distributions for the difference in mean SSVEP RMS amplitude. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is greater than zero. We can see that 55/63 subgroup relationships have $p(\Delta <0 | data)>0.99$.}
\label{fig:eeg_comp}
\end{figure}

\begin{figure}[hptb]
\centering
\includegraphics[width=0.75\linewidth]{../analysis/plots/subgroup_comp_psychophysical.png}
\caption{Posterior distributions for the difference in mean symmetry detection threshold durations. Colour coding of the text indicates the index of the subgroup, while the colour of the filled distribution relates to the conditional probability that the difference in means is smaller than zero. We can see that 43/63 subgroup relationships have $p(\Delta >0 | data)>0.99$.}
\label{fig:durations_comp}
\end{figure}

@@

\textcolor{ForestGreen}{A large literature exists on the \textit{Sustained Posterior Negativity} (SPN), a characteristic negative-going waveform that is known to reflect responses to symmetry and other forms of regularity and structure \citep{makin_2016}. The SPN scales with the proportion of reflection symmetry in displays that contain a mixture of symmetry and noise \cite{makin_formation_2020, palumbo_scaling_2015}, and both reflection, rotation and translation can produce a measurable SPN \cite{makin_2013}. It has recently been demonstrated that a holographic model of regularity \citep{van_der_helm_1996}, can predict both SPN amplitude \citep{makin_2016} and perceptual discrimination performance \citep{nucci_2007} for dot patterns that contain symmetry and other types of regularity. The available evidence suggests that the SPN and our SSVEP measurements are two distinct methods for isolating the same symmetry-related brain response: When observed in the time-domain, the symmetry-selective odd-harmonic responses produce similarly sustained waveforms (see Figure 2), odd-harmonic SSVEP responses can be measured for dot patterns similar to those used to measure the SPN \citep{RN1954}, and the one event-related study on the wallpaper groups also found SPN-like waveforms \citep{RN1959}. Future work should more firmly establish the connection and determine if the SPN can capture similarly precise symmetry responses as the SSVEPs presented here. It would also be worthwhile to ask if and how \textit{W} can computed for our random-noise based wallpaper textures where combinations of symmetries tile the plane.}

@@

\begin{figure}[tbp]
\centering
\includegraphics[width=0.5\linewidth]{../analysis/plots/model_roc_style.pdf}
\caption{This plot shows the proportion of subgroup relationships that satisfy \textcolor{ForestGreen}{$p(\Delta >0 | data) > x$ for the SSVEP data and $p(\Delta <0 | data) > x$ for the psychophysical data}. We can see that if we take $x = 0.95$ as our threshold, the subgroup relationships are preserved in 56/63 = 89\% and \textcolor{ForestGreen}{48/63 = 76\%} of the comparisons for the primary SSVEP and threshold duration datasets, receptively. This compares to the \textcolor{ForestGreen}{32/63= 51\%} and \textcolor{ForestGreen}{22/63 = 35\%} for the SSVEP control datasets.}
\label{fig:durations_rotations}
\end{figure}
